Estimating the number of clusters using Cross Validation

### 2015-06-24 cvclust.tex
minor tweaks to simulation results table
minor tweaks to intro

\begin{document}
\title{Estimating the number of clusters using Cross Validation}
\author{Wei Fu \qquad Patrick O. Perry \\\\ New York University}
\date{}
\maketitle
\begin{abstract}
Many clustering methods, including K-means, require the user to specify the number of clusters as an input parameter. A variety of methods have been devised to choose the number of clusters automatically, but they often rely on strong modeling assumptions. We propose a data-driven approach to estimate the number of clusters based on a novel form of cross-validation. This differs from ordinary cross-validation, because clustering is fundamentally an unsupervised learning problem. Simulation and real data analysis results show that our proposed method outperforms existing methods, especially in high-dimensional settings with heavy-tailed data.
\end{abstract}

\section{Introduction}
As a main task of exploratory data analysis, clustering organizes unlabeled observations into groups such that observations in same group are more similar compare to those in different group. Clustering is an important topic in unsupervised learning because it can reveal the internal structure of data through grouping, segment the data through partitioning and summarize data for other purposes such as dimension reduction. It has being widely used in various fields such as psychology, biology, statistics and machine learning including pattern recognition, image segmentation etc.

After being proposed more than $50$ years, $K$-mean remains one of the most
popular and widely used clustering algorithms \citep{jain2010data}. Like many
other clustering methods, $K$-mean requires an input parameter $k$, the number
of clusters, to be specified by the user. Automatically and quantitatively
deciding such parameter is important and yet unsolved problem
\citep{fujita2014non}. Various methods have been proposed to tackle this
difficulty. One ad hoc approach is to explore the relationship between $W_k$
(within-cluster dispersion) and the number of cluster $k$ for a certain
clustering method such as $K$-mean. Since $W_k$ decreases as $k$ increases,
one usually find the ``elbow" of curve $W_k$ \textsl{VS} $k$ as the
appropriate number of clusters. The example on the top row of Figure
\ref{fig:elbow} demonstrates such approach for data with $k=4$, where the
``elbow" point indeed reveals the true number of clusters. This is based on
the idea that under partitioning data set has more impact than over
partitioning data set in terms of $W_k$. However, locating the ``elbow" point
is somewhat subjective and sometimes is not appropriate to select the optimal
$k$. The second example on the bottom row of Figure \ref{fig:elbow} shows a  situation where there is no clear choice of the ``elbow" point -- both $k=2$ and $k=3$ can be viewed as the ``elbow" point. What's more, the true $k=4$ can never be selected as the optimal $k$ using such approach in this case since it can hardly be viewed as the ``elbow" of the curve.

@@

Recently, there are several new proposals to find the $k$ automatically. Gap statistics \citep{tibshirani2001estimating} estimates $k$ by comparing the change in within-cluster dispersion with that expected under an appropriate reference null distribution. Specifically, the graph of $log(W_k)$ is compared with its expectation under an appropriate null reference distribution of the data. The value of $k$ associated with the largest gap between $log(W_k)$ and the reference curve is selected as optimal $k$. \cite{sugar2003finding} proposed an approach which finds the number of clusters based on distortion, a quantity that measures the average distance, per dimension. It's backed by a rigorous theoretical justification based on information-theoretic ideas.   \cite{fraley2002model}'s Model-based method employs the EM algorithm to estimate the parameters in Gaussian mixture model, and select the best model ($k$) using BIC criterion. Stability-based criterion is also proposed to locate the best $k$ by some authors such as \cite{ben2001stability}, \cite{wang2010consistent} and \cite{fang2012selection}. \cite{chiang2010intelligent} provides a nice review of existing methods for finding the right $k$ in published literature.

@@

\begin{center}
{%\footnotesize
\begin{longtable}{lcccccccccccccccc}
\caption{\label{table1} Simulation Results}\\
\toprule
\phantom{Pred.~Strength}
& \multicolumn{16}{c}{Estimated Number of Clusters} \\
\cmidrule(lr){2-17}
Method
  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15
  & \textsc{na} \\
\midrule
\endfirsthead
\multicolumn{17}{c}%
{\tablename\ \thetable\ -- \textit{Continued from previous page}} \\
\toprule
\phantom{Pred.~Strength}
  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
\cmidrule(lr){2-17}
Method
  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15
  & \textsc{na} \\
\midrule
\endhead
%\bottomrule
\multicolumn{17}{r}{\textit{Continued on next page}} \\
\endfoot
\bottomrule
\multicolumn{17}{l}{
{\footnotesize {In Setting~1, ``\textendash'' means the method can not be used
with parameter $k=1$}.}}
\endlastfoot
\input{demo/bench/setting1/results.tex}
\midrule
\input{demo/bench/setting2/results.tex}
\midrule
\input{demo/bench/setting3/results.tex}
\midrule
\input{demo/bench/setting4/results.tex}
\midrule
\input{demo/bench/setting5/results.tex}
\midrule
\input{demo/bench/setting6/results.tex}
\end{longtable}
}
\end{center}

>>>>

\begin{document}
\title{Estimating the number of clusters using Cross Validation}
\author{Wei Fu \qquad Patrick O. Perry \\\\ New York University}
\date{}
\maketitle
\begin{abstract}
Many clustering methods, including $k$-means, require the user to specify the
number of clusters as an input parameter. A variety of methods have been
devised to choose the number of clusters automatically, but they often rely on
strong modeling assumptions. We propose a data-driven approach to estimate the
number of clusters based on a novel form of cross-validation. This differs
from ordinary cross-validation, because clustering is fundamentally an
unsupervised learning problem. Simulation and real data analysis results show
that our proposed method outperforms existing methods, especially in
high-dimensional settings with heavy-tailed data.
\end{abstract}

\section{Introduction}

As a main task of exploratory data analysis, clustering organizes unlabeled
observations into groups such that observations in same group are more similar
compare to those in different group. Clustering is an important topic in
unsupervised learning because it can reveal the internal structure of data
through grouping, segment the data through partitioning and summarize data for
other purposes such as dimension reduction. It has being widely used in
various fields such as psychology, biology, statistics and machine learning
including pattern recognition, image segmentation etc.

After being proposed more than $50$ years, $k$-means remains one of the most
popular and widely used clustering algorithms \citep{jain2010data}. Like many
other clustering methods, $k$-means requires an input parameter $k$, the
number of clusters, to be specified by the user. Automatically and
quantitatively deciding such parameter is important and yet unsolved problem
\citep{fujita2014non}. Various methods have been proposed to tackle this
difficulty. One ad hoc approach is to explore the relationship between $W_k$
(within-cluster dispersion) and the number of cluster $k$ for a certain
clustering method such as $k$-means. Since $W_k$ decreases as $k$ increases,
one usually find the ``elbow" of curve obtain by plotting $W_k$ versus $k$ as
the appropriate number of clusters. The example on the top row of
Figure~\ref{fig:elbow} demonstrates such approach for data with $k=4$, where
the ``elbow" point indeed reveals the true number of clusters. This is based
on the idea that under partitioning data set has more impact than over
partitioning data set in terms of $W_k$. However, locating the ``elbow" point
is somewhat subjective and sometimes is not appropriate to select the optimal
$k$. The second example on the bottom row of Figure \ref{fig:elbow} shows a
situation where there is no clear choice of the ``elbow" point -- both $k=2$
and $k=3$ can be viewed as the ``elbow" point. What's more, the true $k=4$ can
never be selected as the optimal $k$ using such approach in this case since it
can hardly be viewed as the ``elbow" of the curve.

@@

Recently, there are several new proposals to find the $k$ automatically. Gap
statistics \citep{tibshirani2001estimating} estimates $k$ by comparing the
change in within-cluster dispersion with that expected under an appropriate
reference null distribution. Specifically, the graph of $\log(W_k)$ is
compared with its expectation under an appropriate null reference distribution
of the data. The value of $k$ associated with the largest gap between
$\log(W_k)$ and the reference curve is selected as optimal $k$.
\citet{sugar2003finding} proposed an approach which finds the number of
clusters based on distortion, a quantity that measures the average distance,
per dimension. It's backed by a rigorous theoretical justification based on
information-theoretic ideas.   \citet{fraley2002model}'s Model-based method
employs the EM algorithm to estimate the parameters in Gaussian mixture model,
and select the best model ($k$) using BIC criterion. Stability-based criterion
is also proposed to locate the best $k$ by some authors such as
\citet{ben2001stability}, \citet{wang2010consistent} and
\citet{fang2012selection}. \citet{chiang2010intelligent} provides a nice review
of existing methods for finding the right $k$ in published literature.

@@

\begin{center}
\footnotesize
\begin{longtable}{lcccccccccccccccc}
\caption{\label{table1} Simulation Results}\\
\toprule
\phantom{Pred.~Strength}
  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
\cmidrule(lr){2-17}
Method
  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
\midrule
\endfirsthead
\multicolumn{17}{c}%
  {\tablename\ \thetable\ -- \textit{Continued from previous page}} \\
\midrule
\phantom{Pred.~Strength}
  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
\cmidrule(lr){2-17}
Method
  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
\midrule
\endhead
\multicolumn{17}{r}{\textit{Continued on next page}} \\
\endfoot
\bottomrule
\multicolumn{17}{l}{\footnotesize
  \textsuperscript{$\ast$} CH, Hartigan, and Stability are excluded from
  Setting~1 because they cannot select $k=1$.
}
\endlastfoot
\input{demo/bench/setting1/results.tex}
\midrule
\input{demo/bench/setting2/results.tex}
\midrule
\input{demo/bench/setting3/results.tex}
\midrule
\input{demo/bench/setting4/results.tex}
\midrule
\input{demo/bench/setting5/results.tex}
\midrule
\input{demo/bench/setting6/results.tex}
\end{longtable}
\end{center}
